---
output:
  revealjs::revealjs_presentation:
    incremental: false
    self_contained: true
    keep_md: true
    center: true
    theme: white
    css: ../assets/custom.css
    transition: slide
    slide_level: 1
    includes:
      in_header: ../assets/header.html
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 35)
```

<section class="white-font" data-background="#4E2A84">
<h1> R Fundamentals Review</h1>

<p>
Christina Maimone <br>
<a href="mailto:christina.maimone@northwestern.edu">christina.maimone@northwestern.edu</a><br>
<a href="http://www.it.northwestern.edu/research">Research Computing Services</a>
</p>
</section>

---

## Purpose

Quick review of material covered by DataCamp's [Intro to R](https://campus.datacamp.com/courses/free-introduction-to-r) course.

Additional topics from [Part 1]() and [Part 2]() of the Research Computing workshop materials.



---

<section class="white-font" data-background="#4E2A84">
<h1>Let's Start...</h1>
</section>

---

## Arithmetic 

```{r, eval=FALSE}
2 + 2
5 %% 2
3.452 * 6
2^4
```

---

## Comparisons

```{r, eval=FALSE}
1 < 2
TRUE == FALSE
'a' == "Boy"
```

<aside class="notes">
double equals for testing for equality

Note that character vectors/strings can use single or double quotes.
</aside>

---

## Functions

```{r, eval=FALSE}
log(10)
log(16, base=2)
```

[`log` help page](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Log.html)

<aside class="notes">
Functions are called with `functionName(parameters)`.  

Multiple paramters are comma-separated.  

Paramters can be named. 

For unnamed parameters, the order matters.  

R functions don't change the objects passed to them.
</aside>

---


## Variables

Use the `<-` operator to assign values to variables.  

```{r}
x <- 4
```

`=` also works but is bad practice and less common.

---

## Comments

The comment character is `#`.  

It can be used anywhere on a line to comment out the rest of the line.  

<aside class="notes"> 
no multiline comment characters (use RStudio)
</aside>

---

<section class="white-font" data-background="#4E2A84">
<h1>Vectors and Factors</h1>
</section>

---

## Vectors

Only one type of data allowed!

```{r}
x <- c(1,2,3,4,5) # good
x
x <- c(1, "cat", TRUE) # bad
x
```

<aside class="notes">
Vectors are also called arrays.  You can create a vector with the `c()` function.  `c` stands for combine.
</aside>

---

## Vector Indexing

Indices start with 1 

```{r}
x<-c('a', 'b', 'c', 'd', 'e')
x[1]
```

<aside class="notes">
In some other languages, indices start with 0 instead.
</aside>

---

## Vector Indexing

You can take slices of vectors

```{r}
x[1:3]
```

Or exclude values with a negative sign:

```{r}
x[-1]
```

---

## Vector Indexing

You can use a vector of integers or booleans to select from a vector as well:

```{r}
x[x<'c']
x[c(1,3,5)]
```

---

# Vectors: Checking Values

Check if a value is in a vector with `%in%`

```{r}
'c' %in% x
```

Or get the first index position of a value with `match()`

```{r}
match('b', x)
```

<aside class="notes">
Use the `which` function for all indices
</aside>

---

## Vectors with Functions

Some functions will apply to each element of a vector

Others take a vector as a parameter

---

```{r}
x <- 1:5
log(x)
var(x)
```

---

## Factors 

Categorical variables

```{r}
colors<-c("red", "blue", "green", 
          "red", "red", "blue")
colors<-factor(colors)
levels(colors)
```

---

## Factors

Be careful of assignments

```{r}
colors[1]<-"pink"
colors
```

---

To fix, add a level

```{r}
colors<-factor(colors, 
               levels=c(levels(colors), "pink")) 
colors[1]<-"pink"  
colors
```

---

<section class="white-font" data-background="#4E2A84">
<h1>Special Types</h1>
</section>

---

## Missing Data (`NA`)


```{r}
tmp<-c(1, 2, 5, NA, 6, NA)
is.na(tmp)
sum(is.na(tmp))
```

<aside class="notes">
Missing data in R is encoded as `NA`.  You can test for `NA` (`is.na`).  Some functions will ignore `NA` when doing computations.  Others will ignore missing values if you tell them to.  Others will process `NA` and give you a result of `NA`.  

Why does this work?  How can you sum logical values?  This takes advantage of the trick that TRUE=1 and FALSE=0. 
</aside>

---

## `NA` and Functions
  
```{r}
mean(tmp)
```

`mean` results in `NA`.  It has an option to exclude missing:
  
```{r}
mean(tmp, na.rm=TRUE)
```

<aside class="notes">
Different functions handle NA differently
</aside>

---

## `NA` and Functions

`table` behaves differently than `mean`:
  
```{r}
tmp<-c("A", "A", NA, "B", NA, "A")
table(tmp)
```

It excludes `NA` by default.  

---

## `NA` in Tables

You have to tell `table` to include `NA`:
  
```{r}
table(tmp, useNA = "ifany")
```

---

## Other Special Values

```{r}
c()
0/0
1/0
```

---

<section class="white-font" data-background="#4E2A84">
<h1>More Complex Data Structures</h1>
</section>

---

## Lists

```{r}
l1 <- list(1, "a", TRUE, 1+4i)
l1
```

---

## Lists

```{r}
l2 <- list(title = "Research Bazaar", 
           numbers = 1:10, data = TRUE )
l2
```

<aside class="notes">
Lists are a bit like complex vectors.  An element of a list can hold any other object, including another list.  You can keep multi-dimensional and ragged data in R using lists.  
</aside>

---

## Indexing Lists

```{r}
l2[2]
```

<aside class="notes">
Indexing lists is a little different.  `[[1]]` is the first element of the list as whatever type it was.  `[1]` is a subset of the list -- the first element of the list as a list.  You can also access list elements by name.
</aside>

---

## Indexing Lists

```{r}
l2$numbers
l2[[2]]
l2[[2]][1]
```

<aside class="notes">
Indexing lists is a little different.  ``[[1]]`` is the first element of the list as whatever type it was.  ``[1]`` is a subset of the list -- the first element of the list as a list.  You can also access list elements by name.
</aside>

---

## Matrices

```{r}
matrix(c('a', 'b', 'c', 'd'), nrow=2)
```

<aside class="notes">
Matrices in R are two-dimensional arrays.  All values of a matrix must be of the same type.

Matrices are used sparingly in R, primarly for numerical calculations or explicit matrix manipulation.
</aside>

---

## Matrices

```{r}
y<-matrix(1:25, nrow=5, byrow=TRUE)
y
```

---

## Indexing Matrices 

`matrix[row, column]`

```{r}
y[1,1]
y[1,]
y[,1]
```

<aside class="notes">
Elements in a matrix are indexed like `mat[row, col]`.  Omitting a value for row or column will give you the entire column or row, respectively.

Using just a single index will get the element from the specified position, if the matrix were turned into a vector first
</aside>

---

```{r}
y[1:2,3:4]
y[,c(1,4)]
```


---


## Data Frames

List of heterogeneous vectors of the same length

R's key object >>> We'll review in detail


