---
title: "Exercises Part 1"
author: "Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

<!--
To Knit the file with and without answers
rmarkdown::render('part1_exercises_with_answers.Rmd',
                  output_file = 'part1_exercises_with_answers.html',
                  params = list(answers=TRUE))
rmarkdown::render('part1_exercises_with_answers.Rmd',
                  output_file = 'part1_exercises.html',
                  params = list(answers=FALSE))
--->

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
#knitr::opts_knit$set(root.dir=)
```

# Exercise: Make a vector

Start by making a vector, named `x`, with the numbers 1 through 26. 

Multiply the vector by 2.

Give the resulting vector names A through Z.  Hint: see the documentation page for `names` for an example.

Hint: there is a built in vector called `LETTERS`

```{r, echo=answers}
x <- 1:26
x <- x * 2
names(x) <- LETTERS
x
```

# Exercise: Manipulating Vectors

Set the 10th and 13th elements of the `x` vector you made in the last exercise to missing.

Then select every other element, starting with the second, of the vector.  See the help for `seq` to help with this.

Take the mean of the log of the values in `x`.  Exclude missing values.

```{r, echo=answers}
x[c(10, 13)]<-NA
x
```

```{r, echo=answers}
x[seq(2, length(x), 2)]
```

```{r, echo=answers}
mean(log(x), na.rm=TRUE)
```


# Exercise: Looking up Values

The code below generates a vector with randomly selected letters.  Check to see if your first initial is in the vector.  Which index positions is it at?

Make a table of the frequencies of the letters in `random_letters`.  Sort the table with the `sort` function.

Challenge: Replace occurences of the most frequent letter in `random_letters` with an `*` instead (don't worry about any ties).  Do this without hard coding in the value of the most frequent letter (meaning, use variables, don't type the value of the letter).  Hint: you'll probably need to use the `names` function at some point to get the most frequent letter.

```{r}
random_letters<-sample(letters, 50, replace=TRUE)
```


```{asis, echo=answers}
Note that your output may be different than what's below because of the random selection of values.
```


```{r, echo=answers}
'c' %in% random_letters
which(random_letters == 'c')
sort(table(random_letters))
```

```{r}
# save the table results
table_results <- sort(table(random_letters), decreasing=TRUE)

# get the name of the first element
freq_letter <- names(table_results)[1]

# use that to replace values
random_letters[random_letters == freq_letter] <- '*'

random_letters
```

```{asis, echo=answers}
If you searched the internet for an answer, you may have been able to come up with other solutions.
```



# Exercise: Workspace management

Remove the `x` vector you created above from your workspace.  Check to make sure it's gone.


```{r, echo=answers}
rm(x)
ls() #list all objects, check by looking
"x" %in% ls() # is "x" in the output?
```

```{asis, echo=answers}
You can also see the list of objects in the Environment tab in the upper right window in RStudio. 
```


# Exercise: Packages

Install the package fortunes.  Look at the package documentation to figure out what it does and call the main method of the package.


```{r, echo=answers, eval=FALSE}
install.packages("fortunes")
```

```{asis, echo=answers}
Or use the Packages tab in the bottom right in RStudio to install the package via the GUI.  To load a package with RStudio, check the box next to it in the Packages tab.  To read the documentation, click on the package name in the packages list.
```


```{r, echo=answers}
library(fortunes) # don't forget to load it!
fortune()
```




# Challenge Exercise: Working with Strings

Part 1: Create a variable called `age` with your age.  Using that variable, output your age as part of a sentence like "My age is 25."  Hint: look up the `paste` function.

Part 2: Concatenate all of the letters in `LETTERS` together into a single object, with a "-" between each letter.  Hint: you'll need the `collapse` argument of `paste`.  Save this result as a new object `allletters`.

Load the `stringr` package.  Using a function from the `stringr` package (look at the package documentation to find one), split `allletters` on "-" so that you have individual letters again, saved in another object.  What type of object did you get back after splitting the letters?  How can you get this back to a vector?

Part 3 (advanced): Find a function in base R (not the `stringr` package) that also splits strings (you may need to use Google).  Compare the documentation to the function you found above.  What's the difference? Hint: try supplying a vector with different split patterns to each function and see what output you get. 


```{r, echo=answers}
age<-25
paste("My age is", age)
allletters<-paste(LETTERS, collapse = "-")
allletters
library(stringr)
split_letters<-str_split(allletters, "-")
split_letters
typeof(split_letters)
split_letters <- split_letters[[1]]
```

```{asis, echo=answers}
`strsplit` will also split strings.  Besides some small differences in the handling of regular expressions, the main difference from `str_split` from the `stringr` package is that `str_split` is vectorized over the input strings and patterns, while `strsplit` is only vectorized over the input strings.  This means you can use `str_split` to split the same string on multiple patterns, but to do that with `strsplit` you'd have to make multiple calls.  

In most circumstances, you could use either function.  The `stringr` package incorporates some functions that exist in base R because it standardizes how you use a larger set of functions to manipulate strings.

If you don't know what regular expressions are, take a look at ???.  If you work with text data, or really any large data files, they can save you tons of time reformatting files and data.
```



# Challenge Exercise: Packages and Strings

Install the cowsay package.  Combine the output of the fortunes package (which you installed and used in an exercise above -- if not, go back and do that) with the functionality of the cowsay package.  Hint: you will also need to use the `paste` function with the `collapse` argument.  To keep pretty formatting, also use the `capture.output` function.


```{asis, echo=answers}
This exercise is based on a post by Ista Zahn: http://people.fas.harvard.edu/~izahn/posts/useless-but-fun-r-packages/
```


```{r, echo=answers, eval=FALSE}
install.packages("cowsay")
```

```{r, echo=answers}
library(cowsay)

say( # the cowsay command
  paste( # paste together individual lines of output
    capture.output( # captures printed output as character data - each line is a separate element in a vector
      fortune() # prints a fortune to the screen
      ), 
    collapse="\n" # argument to paste that tells it to join vector elements with a newline
    )
  )
```

```{asis, echo=answers}
Or drop `capture.output`:
```

```{r, echo=answers}
say(paste(fortune(), collapse="\n"))
```

```{asis, echo=answers}
To understand the above better, try running the steps from the inside out to see what each stage feeds the next.
```

```{r, echo=answers, eval=FALSE}
fortune()
capture.output(fortune())
paste(capture.output(fortune()), collapse="\n")
paste(fortune(), collapse="\n") # how is this different from above?
```

```{asis, echo=answers}
`fortune()` returns a list.  `paste` will combine the elements together directly, but `capture.output` gets formatted output that includes spacing.
```


# Exercise: Factors

Use the code below to create a vector with randomly selected names of months.  Then turn this vector into a factor.  Create a table of the value frequencies in the vector with the `table` function.  

Then remake the factor by explicitly setting ordered levels.  Hint: `month.abb` is a built in vector that lists the months in order.

Make a table again.  What's the difference?

```{r}
months<-sample(month.abb, size=40, replace=TRUE)
```


```{r, echo=answers}
months<-sample(month.abb, size=40, replace=TRUE)
months<-factor(months)
table(months) # months should be alphabetical
months<-factor(months, levels=month.abb, ordered=TRUE)
table(months) # months should be in order
```






